### 2019-12-02 paper.rmd
rework front end and move content to notes

---
title             : "Doing Latent Profile Analysis in a more rigorous and reproducible way: A tutorial using the tidyLPA R package"
shorttitle        : "Title"

@@

After describing the method and some examples of its use, we provide a tutorial
for carrying out LPA in the context of a freely-available, open-source
statistical software package we created for R (R Core Team, 2019), tidyLPA.
Finally, we offer some informed recommendations for researchers aiming to use
LPA in their applied work, and conclude with reflections on the role of
statistical software--especially software that is freely-available, open-source,
and highly-performant--in the psychological sciences.

@@

A note of caution is warranted about LPA in the context of their potential.
Bauer (2007) notes that many samples of data can be usefully broken down into
profiles, and that the addition of profiles will likely be suggested for reasons
other than the samples coming from more than one distribution (i.e., due to
non-normality in the variables measured). Bauer also cautions that profiles
should not be reified; that profiles do not necessarily exist outside of the
analysis that they should be interpreted more as useful interpretative devices.
These cautions suggest that, in general, parsimony, interpretability, and a
general sense that the profiles are not necessarily real, but are rather helpful
analytic tools, should be both priorities for the analyst and the reader of
studies using this approach.

@@

SPSS is a common tool to carry out cluster analyses (but not LPA). While somewhat
straightforward to carry out, particularly in SPSS’s graphical user interface
(GUI), there are some challenges to use of this approach. The GUI in SPSS can be
challenging, even for the most able analyst, to be able to document every step
with syntax, and so reproducing the entire analysis efficiently can be a
challenge, both for the analyst exploring various solutions and for the reviewer
looking to replicate this work. Additionally, SPSS is commercial software (and
is expensive), and so analysts without the software cannot carry out this
analysis.

Another common tool is MPlus (Muthen & Muthen, 20xx). MPlus is a commercial tool
that provides functionality for many latent variable (and multi-level) models.
We will speak more to MPlus in the section on tidyLPA, as our software provides
an interface to both it and an open-source tool.

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
this framework, the term “mixture component” has a similar meaning to a profile.
While much more constraining than the latent variable modeling framework, the
approach is often similar or the same: the EM algorithm is used to (aim to)
obtain the maximum likelihood estimates for the parameters being estimated. Like
in the latent variable modeling framework, different models can be specified.

In addition to following the same general approach, using tools that are
designed for Gaussian mixture modeling have other benefits, some
efficiency-related (see RMixMod, which uses compiled C++ code) and others in
terms of ease-of-use (i.e., the plot methods built-in to RMixMod, mclust, and
other tools). However, they also have some drawbacks, in that it can be
difficult to translate between the model specifications, which are often
described in terms of the geometric properties of the multivariate distributions
being estimated (i.e., “spherical, equal volume”), rather than in terms of
whether and how the means, variances, and covariances are estimated. They also
may use different default settings (than those encountered in MPlus) in terms of
the expectation-maximization algorithm, which can make comparing results across
tools challenging.

@@

Because of the limitations in other tools, we set out to develop a tool that a) 
provided sensible defaults and were easy to use, but provided the option to access and modify all of the
inputs to the model (i.e., low barrier, high ceiling), b) interfaced to existing
tools,, and are able to translate between what existing tools are capable of and
what researchers and analysts carrying-out person-oriented analyses would like
to specify, c) made it easy to carry-out fully-reproducible analyses and d) were
well-documented.

@@

A simple summary of the analysis is printed to the console (and its posterior
probability). The resulting object can be further passed down a pipeline to
other functions, such as `plot`, `compare_solutions`, `get_data`, `get_fit`,
etc. This is the "tidy" part, in that the function can be embedded in a tidy
analysis pipeline.

>>>>

---
title             : "Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package"
shorttitle        : "tidyLPA"

@@

After describing the method and some examples of its use, we provide a tutorial
for carrying out LPA in the context of a freely-available, open-source
statistical software package we created for R (R Core Team, 2019), tidyLPA.
Finally, we offer some ideas about best practices and informed recommendations
for researchers aiming to use LPA in their applied work, and conclude with
reflections on the role of statistical software--especially software that is
freely-available, open-source, and highly-performant--in the psychological
sciences.

@@

<!-- I think we move this to the discussion in light of best practices -->

A note of caution is warranted about LPA in the context of their characteristics
and strengths. Bauer (2007) notes that many samples of data can be usefully
broken down into profiles, and that the addition of profiles will likely be
suggested for reasons other than the samples coming from more than one
distribution (i.e., due to non-normality in the variables measured). Bauer also
cautions that profiles should not be reified; that profiles do not necessarily
exist outside of the analysis that they should be interpreted more as useful
interpretative devices. These cautions suggest that, in general, parsimony,
interpretability, and a general sense that the profiles are not necessarily
real, but are rather helpful analytic tools, should be both priorities for the
analyst and the reader of studies using this approach.

@@

SPSS is a common tool to carry out cluster analyses (but not LPA). While
somewhat straightforward to carry out, particularly in SPSS’s graphical user
interface (GUI), there are some challenges to use of this approach. The GUI in
SPSS can be challenging, even for the most able analyst, to be able to document
every step with syntax, and so reproducing the entire analysis efficiently can
be a challenge, both for the analyst exploring various solutions and for the
reviewer looking to replicate this work. Additionally, SPSS is commercial
software (and is expensive), and so analysts access to it cannot carry out this
analysis.

Another common tool is MPlus (Muthen & Muthen, 2019). MPlus is a commercial tool
that provides functionality for many latent variable (and multi-level) models.
We will speak more to MPlus in the section on tidyLPA, as our software provides
an interface to both it and an open-source tool.

In R, a number of tools *can* be used to carry out LPA. OpenMx can be used for
this purpose (and to specify almost any model possible to specify within a
latent variable modeling approach). However, while OpenMx is very flexible, it
can also be challenging to use. Other tools in R allow for estimating Gaussian
mixture models, or models of multivariate Gaussian (or normal) distributions. In
addition to following the same general approach, using tools that are designed
for Gaussian mixture modeling have other benefits, some efficiency-related (see
RMixMod, which uses compiled C++ code) and others in terms of ease-of-use (i.e.,
the plot methods built-in to RMixMod, mclust, and other tools). However, they
also have some drawbacks, in that it can be difficult to translate between the
model specifications, which are often described in terms of the geometric
properties of the multivariate distributions being estimated (i.e., “spherical,
equal volume”), rather than in terms of whether and how the means, variances,
and covariances are estimated. They also may use different default settings
(than those encountered in MPlus) in terms of the expectation-maximization
algorithm, which can make comparing results across tools challenging.

@@

Because of the limitations in other tools, we set out to develop a tool that a)
provided sensible defaults and were easy to use, but provided the option to
access and modify all of the inputs to the model (i.e., low barrier, high
ceiling), b) interfaced to existing tools, and are able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify, c) made it easy to
carry-out fully-reproducible analyses and d) were well-documented.

@@

A simple summary of the analysis is printed to the console. 
The resulting object can be further passed down a pipeline to
other functions, such as `compare_solutions`, `get_data`, `get_fit`,
etc. This is the "tidy" part, in that the function can be embedded in a tidy
analysis pipeline.
